The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2011-10-21 smurf_paper.tex
more work on intro, re-organization of early sections

There are numerous ways to attack this problem, most of them evolving
out of techniques that were developed to produce maps of the
large-scale anisitropies in the Cosmic Microwave Backgroubd.  Styles
of map-making: maximum likelihood \citep[e.g.,][]{patanchon2008};
iterative -- see \citet{johnstone2000} implementation of
\citet{wright1996} to remove chop, \citet{kovacs2008} for SHARC-2
etc.; de-correlation using PCA analysis
\citep[e.g.][]{laurent2005,scott2008,aguirre2010}.

\scuba\ is presently unique in terms of the shear volume of data that
it produces on a nightly basis, and the fact that it is observing a
wide variety of science targets for multiple users (unlike dedicated
CMB experiments, for example). For this reason, SMURF has been
designed to use a less accurate (primarily on large angular scales),
though faster and less memory-intensive iterative approach that
attempts to model and remove most of the correlated signal components,
and then regrid the residual signals. This method is similar in spirit
to the Comprehensive Reduction Utility for SHARC-2
\citep[CRUSH,][]{kovacs2008}, and the pipeline developed for the
Bolocam Galactic Plane Survey \citep{aguirre2010}. While most of the
analysis described here applies equally to all of the data taken
to-date with \scuba, we note that the SCUBA-2 Shared-Risks Observing
period (S2SRO) which took place from 2010 February 11 to 2010 March 25
used only a single subarray at each of the two observing wavelengths,
and had considerably different noise properties and poorer performance
than the fully-commissioned instrument in 2011.

%------------------------------------------------------------------------------
\section{Data Properties}
\label{sec:data}
%------------------------------------------------------------------------------

My thought is that this section would give the basics about how
SCUBA-2 works, and then more specific details for the S2SRO
observations.

%--------------------------------------------------
\subsection{How \scuba\ takes measurements}
%--------------------------------------------------

@@

The \scuba\ bolometer temperatures are read-out using superconducting
transition-edge sensors (TESs), with critical temperatures
$T_\mathrm{C}$ (...typical value...). The range about $T_\mathrm{C}$
in which the transition from a superconductor to a normal material
occurs is extremely small (...typical value...?). If the devices are
biased to a temperature within this transition on large time scales,
then tiny temperature fluctuations on short time-scales will produce
significant changes in the impedence. The entire SCUBA-2 focal plane
is therefore held at a \emph{base temperature} somewhat below
$T_\mathrm{C}$. The TESs then have a small electrical voltage bias,
and individual pixel heaters (using a current bias to apply the
desired Ohmic heating) are used to place the detectors in the
transition. Only a small detector bias voltage is required since the
heaters compensate for most of the variations in the base temperature,
and other ambient loads (like the atmosphere).  Small temperature (and
therefore impedence) fluctuations produce a changing current through
the TESs. The resulting changing magnetic fields are then amplified
using chains of superconducting quantum interference devices (SQUIDs),
and the larger output current is finally digitized. Each detector has
its own SQUID for the first-stage of the amplification, but the
remaining stages occur within a common chain of SQUIDs for each column
of 40 detectors. Each of the 32 rows is then read-out in sequence that
occurs at a frequency of ...kHz?... This multiplexing scheme greatly
reduces the number of readout and bias lines to each detector, making
such large-format bolometer array designs feasible. It should also be
noted that there is an additional 33rd row of SQUID reasouds that are
not connected to TES. These ``dark SQUIDs'' are used to track
non-thermal noise sources that are common to each column's amplifier
chain.

The focal planes at each of 450 and 850\,\micron\ are populated with
four of these $40 \times 32$ subarrays (referred to as s4a--s4d, and
s8a--s8d respectively), and simultaneously image approximately the
same region of the sky by means of a beam-splitter.

The camera has been designed so that the absolute response of each
detector can be established by ramping currents through the individual
pixel heaters to produce Joule heating power, $P_\mathrm{J} =
\epsilon I^2R$, where $I$ is the known current, $R$ is the heater
resistance, and $\epsilon$ is the efficiency with which power is
transferred to the bolometer. In order to calibrate the total factor
$\epsilon R$, SCUBA-2 posesses an internal flatfield source.

@@

@@

The scan strategy is designed, primarily, to mitigate the effects of
$1/f$ drift. The simplest scan one might use is a raster pattern, or a
series of parallel passes across this field. However, such a scan
yields extremely poor performace for structures that are transverse to
the scan direction, since a long time will have ellapsed between each
parallel step, and the signal will therefore be totally dominated by
the $1/f$ noise. In addition, other noise features may appear in the
data periodically. With such a regularly spaced periodic scan
strategy, these noise features can easily conspire with the scan
pattern to produce similarly periodic noise structures in the map, or
\emph{scan-synchronous} noise. It has therefore been recognized for
some time that scan strategies that visit each region of the map on as
many different time scales, and scan angles as possible offer the best
protection against large-scale noise (...cite some people...); such
strategies are said to offer good \emph{cross-linking}. As a tradeoff
between truly random scan patterns and feasibility of implementation
taking into consideration both mechanical and scheduling constraints,
SCUBA-2 presently has three primary scan modes: PONG (cite Borys et
al.), Lissajous, and Daisy... (more citations? Kovacs et al. has a
nice description... also CMB literature).

%-------------------------------------------------
\subsection{\scuba\ shared-risk observations}
%-------------------------------------------------

The S2SRO period, as mentioned above, took place during February and
March 2010. For the purpose of this paper, we will consider data taken
as early as 2009 December 3, at which point the instrument was being
commissioned in a similar configuration to that of the S2SRO
period. At this time, each of the 450 and 850\,\micron\ focal planes
were populated with single subarrays, s4a and s8d respectively. There
were two main mechanical problems to contend with during this
period. First, the fridge was not able to cool the focal plane to the
desired base temperature of ... which affected stability, particularly
at 850\,\micron. Second, there were problems with the subarray
fabrication process which led to large numbers of dead pixels, as well
as large-scale variations in $T_\mathrm{C}$ and the thermal
conductivity, $G$. These latter issues made it difficult to
simulataneously lock all of the bolometers, and in addition many of
the bolometers that were locked did not achieve their minimum
theoretical noise performance, since it is not possible to adjust the
individual temperature biases and place each device into the optimal
region of the superconducting transition.

\subsubsection{Time series}

@@

Both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, from $-0.04$\,pW to $+0.08$, are larger
than the $-0.03$\,pW to $+0.05$\,pW fluctuations at 850\,\micron. Such
behaviour might be expected if there is a comparable varying thermal
load from the fridge at each wavelength, but a larger contribution of
astmospheric variations through the 450\,\micron\ bandpass filters. We
also note that there is no obvious strong correlation between the
low-frequency signal structure at either wavelength at the telescope
motion.

The low-frequency signal component of the bolometer signals is also
highly correlated amongst bolometers in the same subarray. We have
calculate a common-mode signal, $\mathbf{c}(t)$ as the straight
average time series of all the working bolometers. We then fit the
amplitude of $\mathbf{c}(t)$ at each wavelength to the time series
shown in Fig.~\ref{fig:bolos_mix} and remove it, yielding the grey
residual signals. These residuals are quite flat, although still with
noticeable variations. The white noise is also apparent, and larger at
450\,\micron\ as one would expect from the larger loading compared to
850\,\micron.

\subsubsection{Power spectral densities}
\label{sec:psd}

@@

The dotted coloured lines show the PSDs for raw, though flatfielded
data. At each wavelength, there is a clear $1/f$ knee at a few Hz,
followed by a predominantly white spectrum punctuated by line features
above $\gsim 30$\,Hz, and finally roll-off caused by the anti-aliasing
filter above $\gsim 70$\,Hz. As indicated in the previous section, the
correlation between the low-frequency components of the different
bolometer signals is large. The solid black lines in
Fig.~\ref{fig:pspec} indicate the PSDs of the common-modes
$\mathbf{c}(t)$ at each wavelength, which reproduce most of the
low-frequency structures, as well as the high-frequency line
features. The $\mathbf{c}(t)$ otherwise drop substantially below the
individual bolometer PSDs at high-frequency as expected if the
bolometers are dominated by un-correlated white noise. Finally, we
note that the amplitudes and slopes of the $\mathbf{c}(t)$ are
similar at both 450 and 850\,\micron. However, since the
450\,\micron\ white noise is larger than at 850\,\micron\ (shown
approximately by the horizontal dotted lines), the $1/f$ knee occurs
at a \emph{lower} frequency at 450\,\micron.

@@

\subsubsection{Principal component analysis}
\label{sec:pca}

@@

\subsubsection{Magnetic field pickup}
\label{sec:magpickup}

>>>>

There are numerous ways to attack this problem, most of them evolving
out of techniques that were developed to produce maps of the
large-scale anisitropies in the Cosmic Microwave Background (CMB).  Styles
of map-making: maximum likelihood \citep[e.g.,][]{patanchon2008};
iterative -- see \citet{johnstone2000} implementation of
\citet{wright1996} to remove chop, \citet{kovacs2008} for SHARC-2
etc.; de-correlation using PCA analysis
\citep[e.g.][]{laurent2005,scott2008,aguirre2010}.

\scuba\ is presently unique in terms of the shear volume of data that
it produces on a nightly basis, and the fact that it is observing a
wide variety of science targets for multiple users (unlike dedicated
CMB experiments, for example). For this reason, SMURF has been
designed to use a faster, and less memory-intensive algorithm than
CMB-style maximum likelihood map-makers, although it is less accurate
at reproducing large-scale structures. The basic procedure is to
iteratively model and remove most of the bolometer-bolometer
correlated signal components and independent low-frequency noise, and
then regrid the residual signals. This method is similar in spirit to
the Comprehensive Reduction Utility for SHARC-2
\citep[CRUSH,][]{kovacs2008}, and the pipeline developed for the
Bolocam Galactic Plane Survey \citep{aguirre2010}. While most of the
analysis described here applies equally to all of the data taken
to-date with \scuba, we note that the SCUBA-2 Shared-Risks Observing
period (S2SRO) which took place from 2010 February 11 to 2010 March 25
used only a single subarray at each of the two observing wavelengths,
and had considerably different noise properties and poorer performance
than the fully-commissioned instrument in 2011.

%------------------------------------------------------------------------------
\section{\scuba\ data properties}
\label{sec:data}
%------------------------------------------------------------------------------

In this section we summarize how \scuba\ bolometers detect temperature
variations (Section~\ref{sec:bolos}), typical \scuba\ scan strategies
(Section~\ref{sec:scan}).

%--------------------------------------------------
\subsection{How the \scuba\ bolometers work}
\label{sec:bolos}
%--------------------------------------------------

@@

The \scuba\ bolometer temperatures are read-out using superconducting
transition-edge sensors (TESs), with critical temperatures
$T_\mathrm{C}$ (...typical value...). The range about $T_\mathrm{C}$
in which the transition from a superconductor to a normal material
occurs is extremely small (...typical value...?). If the devices are
biased to a temperature within this transition on large time scales,
then tiny temperature fluctuations on short time-scales will produce
significant changes in the impedence. The entire SCUBA-2 focal plane
is therefore held at a \emph{base temperature} somewhat below
$T_\mathrm{C}$. The TESs then have a small electrical voltage biases,
and individual pixel heaters (using a current bias to apply the
desired Joule heating) are used to place the detectors in the
transition. Only a small detector bias voltage is required since the
heaters compensate for most of the variations in the base temperature,
and other ambient loads (like the atmosphere).  Small temperature (and
therefore impedence) fluctuations produce a changing current through
the TESs. The resulting changing magnetic fields are then amplified
using chains of superconducting quantum interference devices (SQUIDs),
and the larger output current is finally digitized. Each detector has
its own SQUID for the first-stage of the amplification, but the
remaining stages occur within a common chain of SQUIDs for each column
of 40 detectors. Each of the 32 rows is then read-out in sequence at a
row-visit frequency of ...kHz?... This multiplexing scheme greatly
reduces the number of readout and bias lines to each detector, making
such large-format bolometer array designs feasible. It should also be
noted that there is an additional 33rd row of SQUID reasouds that are
not connected to TES. These ``dark SQUIDs'' are used to track
non-thermal noise sources that are common to each column's amplifier
chain.

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. Since there
are variations in $T_\mathrm{C}$ across each subarray due to
manufacturing difficulties, considerable effort has gone into
dynamically applying position-dependent biases as well as an over-all
focal plane bias to make as many detectors useable as possible
\citep{holland2011}. In practice, the response of a given detector can
therefore vary somewhat from observation to observation depending on
how the bias and amplification parameters were set up. To track these
changes, at the start of each observation, the current is ramped
through the bolometer heaters such that $dI/dP$ can be measured. Note
that the bolometers may have a slightly different response to heater
power due to conductance as compared to the response to optical power,
so additioanl efficiency corrections must also be obtained. While not
discussed further in this paper, details may be found in
\citet{dempsey2011}. We will refer to measurements of $dI/dP$ across
the focal plane as the \emph{flatfield}. Fig.~\ref{fig:sensitivities}
shows the flatfield for each focal plane.

@@

%--------------------------------------------------
\subsection{Scan strategies}
\label{sec:scan}
%--------------------------------------------------

@@

The scan strategy is designed, primarily, to mitigate the effects of
$1/f$ drift. The simplest scan one might use is a raster pattern, or a
series of parallel passes across this field. However, such a scan
yields extremely poor performace for structures that are transverse to
the scan direction, since a long time will have ellapsed between each
parallel step, and the signal will therefore be totally dominated by
the $1/f$ noise. In addition, other noise features may appear in the
data periodically. With such a regularly spaced periodic scan
strategy, these noise features can easily conspire with the scan
pattern to produce similarly periodic noise structures in the map, or
\emph{scan-synchronous} noise. It has therefore been recognized that
scan strategies that visit each region of the map on as many different
time scales, and scan angles as possible offer the best protection
against large-scale noise (...cite some people...); such strategies
are said to offer good \emph{cross-linking}. As a tradeoff between
truly random scan patterns and feasibility of implementation taking
into consideration both mechanical and scheduling constraints, SCUBA-2
presently has three primary scan modes: PONG (cite Borys et al.),
Lissajous, and Daisy... (more citations? Kovacs et al. has a nice
description... also CMB literature).

%-------------------------------------------------
\subsection{Typical bolometer signals}
%-------------------------------------------------

@@

Both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, from $-0.04$\,pW to $+0.08$, are larger
than the $-0.03$\,pW to $+0.05$\,pW fluctuations at 850\,\micron. Such
behaviour might be expected if there is a comparable varying thermal
load from the fridge at each wavelength, but a larger contribution of
atmospheric variations through the 450\,\micron\ bandpass filters. We
also note that there is no obvious strong correlation between the
low-frequency signal structure, at either wavelength, with the
telescope motion.

The low-frequency signal component of the bolometer signals is also
highly correlated amongst bolometers in the same subarray. We have
calculated a common-mode signal, $\mathbf{c}(t)$, as the average time
series of all the working bolometers. We then fit the amplitude of
$\mathbf{c}(t)$ at each wavelength to the signals shown in
Fig.~\ref{fig:bolos_mix} and remove it, yielding the grey residual
signals. These residuals are quite flat, although still with
noticeable variations. The white noise is also apparent, and larger at
450\,\micron\ as one would expect from the larger backgrounds compared
to 850\,\micron.

@@

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flatfielded data. At each wavelength, there is a clear
$1/f$ knee at a few Hz, followed by a predominantly white spectrum
punctuated by line features above $\gsim 30$\,Hz, and finally roll-off
caused by the anti-aliasing filter above $\gsim 70$\,Hz. As indicated
in the previous section, the correlation between the low-frequency
components of the different bolometer signals is large. The solid
black lines in Fig.~\ref{fig:pspec} indicate the PSDs of the
common-modes $\mathbf{c}(t)$ at each wavelength, which reproduce most
of the low-frequency structures, as well as the high-frequency line
features. The $\mathbf{c}(t)$ otherwise drop substantially below the
individual bolometer PSDs at high-frequency as expected if the
bolometers are dominated by un-correlated white noise. Finally, we
note that the amplitudes and slopes of the $\mathbf{c}(t)$ are similar
at both 450 and 850\,\micron. However, since the 450\,\micron\ white
noise is larger than at 850\,\micron\ (shown approximately by the
horizontal dotted lines), the $1/f$ knee occurs at a \emph{lower}
frequency at 450\,\micron.

@@

\subsection{Magnetic field pickup}
\label{sec:magpickup}

%-------------------------------------------------
\subsection{Comparison between \scuba\ shared-risk observations and the 2011 re-commissioned instrument}
%-------------------------------------------------

The S2SRO period took place during February and March 2010. For the
purpose of this paper, we will consider data taken as early as 2009
December 3, at which point the instrument was being commissioned in a
similar configuration to that of the S2SRO period. At this time, each
of the 450 and 850\,\micron\ focal planes were populated with single
subarrays, s4a and s8d respectively. There were two main mechanical
problems to contend with during this period. First, the fridge was not
able to cool the focal plane to the desired base temperature of
... which affected stability, particularly at 850\,\micron. Second,
there were problems with the subarray fabrication process which led to
large numbers of dead pixels, as well as large-scale variations in
$T_\mathrm{C}$ and the thermal conductivity, $G$. These latter issues
made it difficult to simulataneously place all of the TESs into the
optimal region of the superconducting transition, leading to
significantly poorer sensitivities.

%-------------------------------------------------
\subsection{Principal component analysis}
\label{sec:pca}
%-------------------------------------------------

@@
